Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Section 5.3

\subsection{A first realization}
\label{sec:keil-thiemann}
\unsure{Subsection title?}

@@

\subsection{Monitoring properties}
\label{sec:will-morr-wadl}
\unsure{Subsection title?}

As mentioned in Section~\ref{sec:issues-sem}, Williams, Morris, and Wadler,
downright exclude user-defined contracts from the picture. This choice gives
them more freedom in the quest of a more uniform operational semantics. While
this can be a sensible choice (in a gradual typing context e.g.), this is a show
stopper for our initial use-case of the Nickel configuration language, the main
motivation being to use contracts as customizable validation functions.

On the semantic side, one of the so-called sound monitoring properties
linking the operational semantics to the denotational semantics (supposed to
capture the intuitive interpretation) is at odds with the given interpretation.
% On top of that, a fundamental problem arises on said paper,
% symptom of the difficulty of establishing good enough, and yet
% simple semantics for union and intersection dynamic checks.
% One of their main results, is the fact that their calculus obeys
% all of their expected monitoring properties for higher ordered, union
% and intersection contracts.

For a contract $A$, $\llbracket A \rrbracket^+$ defines the set of terms that
satisfy $A$. For example,

\begin{lstlisting}[language=nickel,frame=none,numbers=none]
    if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
\end{lstlisting}

For higher-order contracts, the notion has two facets. A \nickel{Num -> Str}
contract can be violated in two ways: by the function returning a value that is
not a string, or by the caller providing an argument that is not a number.  The
former is embodied by the same notion as before: $f \in \llbracket Num \to Str
\rrbracket^+$ if whenever $f$ receives a number, it returns a string.  One
important contribution of~\cite{KeilThiemannUnionIntersection}, upon
which~\cite{RootCauseOfBlame} builds, is to extend the notion of contract
satisfaction to \emph{contexts}, embodying the caller's
obligations. A context $K$ is a term with a hole to be filled with a term (Figure~\ref{fig:valid-contexts} has examples). Then, $K \in \llbracket Num \to
Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})
if it only applies the hole to arguments satisfying the \nickel{Num} contract.

These interpretations $\llbracket . \rrbracket^+$ and $\llbracket .
\rrbracket^-$ then need to be extended to the setting of unions and
intersections. In particular, one of the definitions of~\cite{RootCauseOfBlame}
is the following:

\begin{figure}[h]
$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
\caption{Sound monitoring property for intersection from~\cite{RootCauseOfBlame}}
\label{fig:wmw-semantics}
\end{figure}

@@

Although valid, this interpretation is weaker than what~\cite{RootCauseOfBlame}
actually describes it to be and than the semantics given
in~\cite{KeilThiemannUnionIntersection}. Consider the two contexts presented on
Figure~\ref{fig:valid-contexts}. The first one is a context satisfying
\nickel{Num -> Num}, applying the hole to a number. Similarly, the second context
from the same figure satisfies \nickel{Bool -> Bool}.

@@

Now, combining these two context as in Figure~\ref{fig:invalid-context} gives a context
that doesn't satisfy \nickel{Num -> Num} nor \nickel{Bool -> Bool}.  According
to the semantics of~\cite{RootCauseOfBlame}, it thus doesn't satisfy \nickel{Num
-> Num @& Bool -> Bool}. Yet, our intuition tells us that this example is a
perfectly valid context for this contract: an overloaded function satisfying
both \nickel{Num -> Num} and \nickel{Bool -> Bool} can safely be applied to a
number and to a boolean in a row. And indeed, the denotational semantics given
in~\cite{KeilThiemannUnionIntersection} does accept this context as valid.

@@

The issue is not as bad as it sounds: the system presented
in~\cite{RootCauseOfBlame} actually corresponds to the wished, stronger
semantics in practice. The specification is just not exactly the right one, but
this can probably be fixed along the lines
of~\cite{KeilThiemannUnionIntersection}, where the notion of context used to
define satisfaction is carefully restricted (\emph{elimination contexts}). While
this kind of oversight doesn't lower the quality of the work
of~\cite{RootCauseOfBlame} by any measure, it illustrates that contract
satisfaction in presence of unions, intersections and functions is a tricky
notion. This difficulty contrasts with the standard framework of higher-order
contracts where satisfaction is much more straightforward.
% Next, we can combine these two contexts into one, that essentially
% has one single missing value, that will be used independently in both contexts
% from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
% This context is not a valid \nickel{Num -> Num} context, since it will apply
% the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
% In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
% Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
% shown on Figure \ref{fig:wmw-semantics}, it is not a valid
% \nickel{Num -> Num @& Bool -> Bool} context.

% However, both our intuition, as the most common use case of intersections
% (that is, overloading), tells us that this should work.
% While this problem does not invalidate their work (in fact, the system they present
% handles these cases correctly), it shows that existing research is still not on
% good shape to handle the growing usecases being developed for
% unions and intersections around dynamic languages.

% $$ K ::= Id | K \bullet \square N | K \bullet V \square | K \bullet \square @^pA  $$

% Keil and Thiemann did noticed this problem, and their contract satisfaction
% rules only apply to elimination context, a particular kind of context
% where the hole is either being applied to an argument
% ($\square N$) or is used as an argument for a primitive
% operation ($O(\overrightarrow{\text{V}}, \square, \overrightarrow{\text{N}})$).

% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
\resolved{(Yann) Maybe adding a concrete example of this (it can be the same term but with
    concrete types like Number and String and simple contexts like application
    to a dumb argument) is sufficient to make the point}

>>>>

\subsection{A first realization}
\label{sec:keil-thiemann}

@@

\subsection{Monitoring properties}
\label{sec:will-morr-wadl}

Another realization of the coinductive semantics described in
Section~\ref{sec:coinductive-sem}, is~\cite{RootCauseOfBlame}, which
aims at simplifying the algorithmic system proposed
by~\cite{KeilThiemannUnionIntersection} and described in
Section~\ref{sec:keil-thiemann}.

A key ingredient of~\cite{RootCauseOfBlame} is disallow user-defined
contracts. This choice gives the authors more freedom in the quest of
a more uniform operational semantics. This is sensible trade-off in
the context of gradual typing à la TypeScript: the problem is to match
contract with static types, and user-defined contracts don't have a
static type equivalent. On the other hand, the cost would probably not
be worth it for a configuration language like Nickel.

% Some of this we may want to move to Section 5.1
%
% For a contract $A$, $\llbracket A \rrbracket^+$ defines
% the set of terms that satisfy $A$. For example,

% \begin{lstlisting}[language=nickel,frame=none,numbers=none]
%     if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
% \end{lstlisting}

% For higher-order contracts, the notion has two facets. A \nickel{Num -> Str}
% contract can be violated in two ways: by the function returning a value that is
% not a string, or by the caller providing an argument that is not a number.  The
% former is embodied by the same notion as before: $f \in \llbracket Num \to Str
% \rrbracket^+$ if whenever $f$ receives a number, it returns a string.  One
% important contribution of~\cite{KeilThiemannUnionIntersection}, upon
% which~\cite{RootCauseOfBlame} builds, is to extend the notion of contract
% satisfaction to \emph{contexts}, embodying the caller's
% obligations. A context $K$ is a term with a hole to be filled with a term (Figure~\ref{fig:valid-contexts} has examples). Then, $K \in \llbracket Num \to
% Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})
% if it only applies the hole to arguments satisfying the \nickel{Num} contract.

As a means of proving the correctness of their simplified system, the
authors introduce what they call \emph{sound monitoring
  properties}. Here is the sound monitoring property for contexts of
intersection contracts:

$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$

@@

Although sound, this interpretation is weaker than what the
coinductive semantics permit. Consider the two contexts presented on
Figure~\ref{fig:valid-contexts}. The first one is a context satisfying
\nickel{Num -> Num}, applying the hole to a number. Similarly, the second context
from the same figure satisfies \nickel{Bool -> Bool}.

@@

Now, combining these two context as in Figure~\ref{fig:invalid-context} gives a context
that doesn't satisfy \nickel{Num -> Num} nor \hbox{\nickel{Bool -> Bool}.}  According
to the sound monitoring property of intersection, Figure~\ref{fig:invalid-context} thus doesn't satisfy \nickel{Num
-> Num @& Bool -> Bool}.

@@

The consequence is that~\cite{RootCauseOfBlame} doesn't prove their
system complete for the coinductive semantics. It's probably just an
oversight in the proof: we believe their system to be indeed complete
for the coinductive semantics. But it does speak to the intrinsic
complexity of union and intersection contracts: it is remarkably easy
to get details wrong. This difficulty contrasts with the standard
framework of higher-order contracts where satisfaction is much more
straightforward.
